---
title: "Network-Valued Data Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Network-Valued Data Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(nevada)
```

In the framework of Object Oriented Data Analysis (OODA), statistical
tools for the analysis of network-valued data are almost lacking,
whereas data sets composed of networks are more and more increasing. The
package `nevada` allows to test if two samples of networks came from the
same population via two test statistics based on three possible
mathematical representation of a network and on four different types of
distances.

## Introduction

A network $G=(V,E)$ is a complex object consisting of a set $V$ of
vertices whose relationships are represented by the edges in the set
$E$. The package `nevada` contains some functions that allow not only to
perform the test, but also to carry out all the intermediate steps that
lead to the test itself. The theoretical details are contained in
@lovato2020, as well as simulations and application to real data. The
logical path followed in @lovato2020 can be summarized as follow. First
of all, choose a matrix representation for the networks and define some
distances between two networks. Then, define two different test
statistics and introduce the permutation framework to perform the test.
Other ref is @lovato2021.

## Network representation

There are three possible matrix representations for a network. Let $G$
be a network with $N$ nodes.

**Adjacency matrix** A $N$ x $N$ matrix $W$ is an adjacency matrix for
$G$ if element $W_{ij}$ indicates if there is an edge between vertex $i$
and vertex $j$: $$
W_{ij}=
\begin{cases}
w_{i,j}, \quad if \quad (i,j) \in E \quad with \quad weight \quad  w_{i,j}\\
0, \quad if \quad (i,j) \not \in E
\end{cases}
$$

**Laplacian matrix** The Laplacian matrix $L$ of the network $G$ is
defined in the following way: $$
L=D(W)-W
$$ where $D(W)$ is the diagonal matrix whose $\{ i,i \}$ element is the
degree of vertex $i$.

**Modularity matrix** The elements of the modularity matrix $B$ are
given by \begin{equation*}
    B_{ij} = W_{ij} - \frac{d_i d_j}{2m}
\end{equation*} where $d_i$ and $d_j$ are the degree of vertices $i$ and
$j$, respectively, and $m$ is the total number of edges in the network.

The functions `repr_*()`, having as input an igraph object, return the
chosen representation.

## Distances between networks

In the package `nevada` it is possible to choose which distance consider
in the analysis. Let $G$ and $H$ be two networks with $N$ nodes each and
suppose that $X$ and $Y$ are the matrix representations of $G$ and $H$,
respectively. The user can choose between the following distances:

-   **Hamming distance**
    $\rho_H(G,H)=\frac{1}{N(N-1)}\sum_{i \neq j}^N \bigl\arrowvert X_{i,j}-Y_{i,j} \bigr\arrowvert$
-   **Frobenius distance**
    $\rho_F(G,H)=\sum_{i \neq j}^N \bigl ( X_{i,j}-Y_{i,j} \bigr )^2$
-   **Spectral distance**
    $\rho_S(G,H)=\sum_{i \neq j}^N \bigl ( \Lambda^X_{i,j}-\Lambda^Y_{i,j} \bigr )^2$,
    where $\Lambda^X$ and $\Lambda^Y$ are the diagonal matrices with
    eigenvalues on the diagonal given by the spectral decomposition of
    the matrix representations of $G$ and $H$,
-   **Root Euclidean distance**
    $\rho_{RE}(G,H)=\rho_F \bigl (\sqrt{L_X}, \sqrt{L_Y}\bigr )$, where
    $\sqrt S=U\sqrt{\Lambda} U^T$ and
    $(\sqrt\Lambda)_{i,j}=\sqrt{\Lambda_{i,j}}$, where $S=U\Lambda U^T$
    from the spectral decomposition of $S$.

The functions `dist_*()` return the chosen distance between two networks
and moreover request the choice of the matrix representation as input.

## Test statistics

The package `nevada` allows to use two different test statistics. The
first one is given by the distance between the restricted Frechet means
of the two samples. The Frechet mean of a set of objects is a sort of
mean element and this concept of average has been widely used in the
literature. The second test statistic is given by the mean of the
distances between all the elements of the first sample and those of the
second one.

It is possible to obtain the values of the test statistics using the
functions `stat_*()`.

## The permutational framework

The hypothesis is tested in a permutational framework. The null
distribution is obtained by permuting the labels of the subjects of the
two populations. The exhaustive list of all possible permutations is
used if its length is smaller than either the user-defined number of
random permutations requested or the number of permutations required for
achieving a user-defined p-value resolution. Otherwise, the p-value is
estimated via Monte-Carlo simulations.

The function `network_test2p()` returns the value of the statistic for
the original two samples and the p-value of the resulting permutation
test.

## References
