---
title: "Network-Valued Data Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Network-Valued Data Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(nevada)
```

*Network-valued data* are data in which the statistical unit is a network itself. This is the data with which we can make inference on *populations of networks* from *samples of networks*. The [**nevada**](https://astamm.github.io/nevada) package proposes a specific `nvd` class to handle network-valued data. Inference from such samples is made possible though a 4-step procedure:

- Choose a suitable representation of your samples of networks.
- Choose a suitable distance to embed your representation into a nice metric space.
- Choose one or more test statistics to define your alternative hypothesis.
- Compute an empirical permutation-based approximation of the null distribution.

The package focuses for now on the two-sample testing problem and assumes that all networks from both samples share the same node structure.

There are two types of questions that one can ask:

1. Is there a difference between the distributions that generated the two observed samples?
2. Can we localize the differences between the distributions on the node structure?

The [**nevada**](https://astamm.github.io/nevada) package offers a dedicated function for answering each of these two questions:

- [`test2_global()`](https://astamm.github.io/nevada/reference/test2_global.html); for more details, please see @lovato2020,
- [`test2_local()`](https://astamm.github.io/nevada/reference/test2_local.html); for more details, please see @lovato2021.

## The `nvd` class for network-valued data

In [**nevada**](https://astamm.github.io/nevada), network-valued data are stored in an object of class `nvd`, which is basically a list of [**igraph**](https://igraph.org/r/) objects. We provide:

- a constructor [`nvd()`](https://astamm.github.io/nevada/reference/nvd.html) which allows the user to simulate samples of networks using some of the popular models from [**igraph**](https://igraph.org/r/). Currently, one can use:

  - the stochastic block model,
  - the $k$-regular model,
  - the GNP model,
  - the small-world model,
  - the PA model,
  - the Poisson model,
  - the binomial model.

The constructor simulates networks with 25 nodes.

- a function [`as_nvd()`](https://astamm.github.io/nevada/reference/as_nvd.html) to coerce lists of [**igraph**](https://igraph.org/r/) objects into an object of class `nvd`.

## Network representation

There are currently 3 possible matrix representations for a network. Let $G$
be a network with $N$ nodes.

### Adjacency matrix

A $N$ x $N$ matrix $W$ is an *adjacency matrix* for
$G$ if element $W_{ij}$ indicates if there is an edge between vertex $i$
and vertex $j$: $$
W_{ij}=
\begin{cases}
w_{i,j}, & \mbox{if } (i,j) \in E \mbox{ with weight } w_{i,j}\\
0, & \mbox{otherwise.}
\end{cases}
$$

In [**nevada**](https://astamm.github.io/nevada), this representation can be achieved with [`repr_adjacency()`](https://astamm.github.io/nevada/reference/representations.html).

### Laplacian matrix

The *Laplacian matrix* $L$ of the network $G$ is
defined in the following way:

$$ L = D(W) - W, $$ 
where $D(W)$ is the diagonal matrix whose $i$-*th* diagonal element is the
degree of vertex $i$.

In [**nevada**](https://astamm.github.io/nevada), this representation can be achieved with [`repr_laplacian()`](https://astamm.github.io/nevada/reference/representations.html).

### Modularity matrix

The elements of the *modularity matrix* $B$ are
given by

$$ B_{ij} = W_{ij} - \frac{d_i d_j}{2m}, $$
where $d_i$ and $d_j$ are the degrees of vertices $i$ and
$j$ respectively, and $m$ is the total number of edges in the network.

In [**nevada**](https://astamm.github.io/nevada), this representation can be achieved with [`repr_modularity()`](https://astamm.github.io/nevada/reference/representations.html).

### Choosing a representation for an object of class `nvd`

Instead of going through every single network in a sample to make its representation, [**nevada**](https://astamm.github.io/nevada) provides the [`repr_nvd()`](https://astamm.github.io/nevada/reference/repr_nvd.html) function which does exactly that for an object of class `nvd`.

```{r}
x <- nvd(model = "gnp", n = 3)
repr_nvd(x, representation = "laplacian")
```

## Distances between networks

It is possible to choose which distance consider in the analysis. Let $G$ and $H$ be two networks with $N$ nodes each and suppose that $X$ and $Y$ are the matrix representations of $G$ and $H$, respectively. The user can currently choose among 4 distances: Hamming, Frobenius, spectral and root-Euclidean.

### Hamming distance

$$ \rho_H(G,H)=\frac{1}{N(N-1)}\sum_{i \neq j}^N \bigl\arrowvert X_{i,j}-Y_{i,j} \bigr\arrowvert. $$

In [**nevada**](https://astamm.github.io/nevada), this distance can be computed with [`dist_hamming()`](https://astamm.github.io/nevada/reference/distances.html).

### Frobenius distance

$$ \rho_F(G,H) = \left\| X - Y \right\|_F^2 = \sum_{i \neq j}^N \bigl ( X_{i,j}-Y_{i,j} \bigr )^2. $$

In [**nevada**](https://astamm.github.io/nevada), this distance can be computed with [`dist_frobenius()`](https://astamm.github.io/nevada/reference/distances.html).

### Spectral distance

$$ \rho_S(G,H)=\sum_{i \neq j}^N \bigl ( \Lambda^X_{i,j}-\Lambda^Y_{i,j} \bigr )^2, $$
where $\Lambda^X$ and $\Lambda^Y$ are the diagonal matrices with
eigenvalues on the diagonal given by the spectral decomposition of
the matrix representations of $G$ and $H$.

In [**nevada**](https://astamm.github.io/nevada), this distance can be computed with [`dist_spectral()`](https://astamm.github.io/nevada/reference/distances.html).

### Root Euclidean distance

$$ \rho_{RE}(G,H) = \left\| X^{1/2} - Y^{1/2} \right\|_F^2. $$

Note that this distance is not compatible with all matrix representations as it requires that the representation be semi-positive definite.

In [**nevada**](https://astamm.github.io/nevada), this distance can be computed with [`dist_root_euclidean()`](https://astamm.github.io/nevada/reference/distances.html).

### Computing a matrix of pairwise distances for an object of class `nvd`

It alleviates the computational burden of permutation testing to pre-compute the matrix of pairwise distances for samples of networks. This is why [**nevada**](https://astamm.github.io/nevada) provides the convenient [`dist_nvd()`](https://astamm.github.io/nevada/reference/dist_nvd.html) function which does exactly that for an object of class `nvd`.

```{r}
x <- nvd(model = "gnp", n = 3)
dist_nvd(x, representation = "laplacian", distance = "hamming")
```

## Test statistics

The nevada package has been designed to work well with the flipr package, which handles the permutation scheme once suitable representation, distance and test statistics have been chosen. The most efficient way to two-sample testing with network-valued data pertains to use statistics based on inter-point distances, that is pairwise distances between observations.

There has been a number of test statistics along this line that have been proposed in the literature, including ours [@lovato2020]. As these test statistics rely on inter-point distances, they are not specific to network-valued data. As such, they can be found in flipr. We adopt the naming convention that a test statistic function shall start with the prefix `stat_`. All statistics based on inter-point distances are named with the suffix `_ip`.

In nevada, there are also 3 statistics proposed in @chen2018weighted that are based on a similarity graph built on top of the distance matrix:

- [`stat_original_edge_count()`](https://astamm.github.io/nevada/reference/statistics.html),
- [`stat_generalized_edge_count()`](https://astamm.github.io/nevada/reference/statistics.html),
- [`stat_weighted_edge_count()`](https://astamm.github.io/nevada/reference/statistics.html).

There are also Student-like statistics available only for Frobenius distance for which we can easily compute the Fréchet mean. These are:

- [`stat_student_euclidean()`](https://astamm.github.io/nevada/reference/statistics.html),
- [`stat_welch_euclidean()`](https://astamm.github.io/nevada/reference/statistics.html).

Test statistics are passed to the functions `test2_global()` and `test2_local()` via the argument `stats` which accepts a character vector in which:

- statistics from nevada expected to be named without the `stat_` prefix (e.g. `"original_edge_count"` or `"student_euclideam"`).
- statistics from flipr are expected to be named without the `stat_` prefix but adding the `flipr:` prefix (e.g., `"flipr:student_ip"`).
- statistics from any other package **pkg** are expected to be named without the `stat_` prefix but adding the `pkg:` prefix.

```{r}
x <- nvd(model = "gnp", n = 10)
y <- nvd(model = "pa", n = 10)
test2_global(
  x = x, 
  y = y, 
  representation = "laplacian", 
  distance = "frobenius", 
  stats = c("flipr:student_ip", "flipr:fisher_ip")
)
```

The package `nevada` allows to use two different test statistics. The
first one is given by the distance between the restricted Frechet means
of the two samples. The Frechet mean of a set of objects is a sort of
mean element and this concept of average has been widely used in the
literature. The second test statistic is given by the mean of the
distances between all the elements of the first sample and those of the
second one.

It is possible to obtain the values of the test statistics using the
functions `stat_*()`.

## The permutational framework

The hypothesis is tested in a permutational framework. The null
distribution is obtained by permuting the labels of the subjects of the
two populations. The exhaustive list of all possible permutations is
used if its length is smaller than either the user-defined number of
random permutations requested or the number of permutations required for
achieving a user-defined p-value resolution. Otherwise, the p-value is
estimated via Monte-Carlo simulations.

The function `network_test2p()` returns the value of the statistic for
the original two samples and the p-value of the resulting permutation
test.

## References
